---
title: "EDPS 845 Assignment 1"
author: "Tony Albano"
---

## Overview

Demonstrate your mastery of concepts and procedures covered in *An Introduction to R*, chapters 1 through 9. See Appendix A for examples.

Complete this assignment by responding to each question below using your own original code, in addition to answering any questions. Follow Google's [R coding conventions](https://google.github.io/styleguide/Rguide.xml), and use [R markdown](http://rmarkdown.rstudio.com/index.html) formatting for your responses, with example R code written as

```{r}
# Comment within some example code
```

Your file should have an Rmd extension, and you should check that it compiles to HTML using the *Knit* command in RStudio (this runs `knitr::knit()` on the file). Note that you can already *Knit* this file to see how it works.

## Exercises

### Chapter 1

1. Show that you can access the home page for the HTML R help menu, as well as help files for individual functions.

```{r, eval = FALSE}
#This will open html help
help.start()

#Both of the following will open the help file for the solve function 
help.solve
?solve
```

2. Demonstrate correct and incorrect naming of R objects.

```{r}
#Best practice dictates NOT using periods, use _ instead
#names can include a number, but cannot begin with a number
#Every character matters and is case-sensitive

#correct
x1 <- c(1,2,3) #where x1 is the object
X1 <- c(2,3,4) #different from lowercase x1

#incorrect
# 2x <- c(3,4,5)
```

3. How do you list all available objects in the R workspace? And how do you remove individual objects?

```{r}
# to display the objects which are currently stored within R
objects()

# to remove individual objects
rm(x1,X1)
```


### Chapter 2

1. Create and index a vector using a logical, integer, negative integer, and character vector.

```{r}

v1 <- 1:30 #creating a vextor that includes the integers 1, 2,...29, 30

#logical vector index

v1logical <- v1[v1%%2==0] #vector "v1logical" where only the values that are included are those for which the index vector in brackets is TRUE. Here, this would give you only the even values. 

# positive integer index

v1first10 <- v1[1:10] #vector that selects the first ten values of vector v1

#negative integer index

v1last10 <- v1[-(1:20)] #vector that excludes values 1-20 of vector v1

#character vector index

faces <- c(12, 11, 36, 24) #creates a vector of those values
names(faces) <- c("happy", "angry", "neutral", "surprised") #assigns names to identify each component of a vector
stim_list1 <- faces[c("neutral", "surprised")] #now I can create another vector comprised of the numeric mode values, without having to remember the specific numeric values because they have names

```

2. Create a simple data frame containing a numeric variable, character variable, and factor with at least 3 levels. Use the data frame to demonstrate how R codes and handles missing data. 

### Chapter 3

1. Demonstrate how to check and change the modes and attributes of R objects.

```{r}

mode(v1) #returns what the mode is; in this case, "numeric"
length (v1) #returns the length; in this case, 30
attributes(faces) #returns additional, non-instrinsic attributes, like names

as.character(v1) #changes all numeric values in vector v1 to character mode
as.integer(v1) #changes all characters in vector v1 back to integer mode

v2 <- numeric() #creates empty vector of mode numeric
length (v2) <-20 #makes the length 20, all missing values (NA)
v2[5] <- 5.1 #makes the 5th value 5.1

attr(v2, "dim") <- c(5,4) #changes the dimensions to a 5-by-4 matrix

```


### Chapter 4

1. Complete the exercise on page 17, where you calculate 95% confidence intervals for state mean incomes.
```{r}

state <- c("tas", "sa", "qld", "nsw", "nsw", "nt", "wa", "wa","qld", "vic", "nsw", "vic", "qld", "qld", "sa", "tas","sa", "nt", "wa", "vic", "qld", "nsw", "nsw", "wa","sa", "act", "nsw", "vic", "vic", "act")

statef <- factor(state) 

incomes <- c(60, 49, 40, 61, 64, 60, 59, 54, 62, 69, 70, 42, 56,61, 61, 61, 58, 51, 48, 65, 49, 49, 41, 48, 52, 46,59, 46, 58, 43)

incmeans <- tapply(incomes, statef, mean) #apply the mean() function to each group of income by levels of statef

stderr <- function(x) sqrt(var(x)/length(x)) #creates a funciton to calculate standard error

incster <- tapply (incomes, statef, stderr) #apply the newly created stderr function to each group of income by levels of statef

ns <- table(statef) #get n for each state/providence

qt (.025, ns-1) #returns CI with lower tail included for each n

#Extra -- plotting

library("ggplot2")
ggplot(cis,aes(state,mean))+
  geom_point()+
  geom_errorbar(aes(mean - stderr, ymin = mean - stderr))

```


2. In what order are the levels of ordered factors stored in?
```{r}
#They are stored in alphabetical order, or in the order they were specified to factor if they were specified explicitly

ex1 <- c("blue", "red", "green")
ex1f <- factor (ex1) #alphabetical


ex1f2 <- factor(ex1, levels=c("blue", "red", "green"))
levels(ex1f2) #specified order

```

### Chapter 5

1. Create an array with dimensions `c(3, 6, 2)` and use it to demonstrate array indexing.
```{r}


```

2. Create a two dimensional array of 100 integers sampled with replacement from the vector 1:20, and use an index array to set the diagonal values to 0.

3. Construct the same array using the `array()` function and using a vector with the `dim()` set. Prove they are the same.

4. Use your array from 5.1, and another with the same dimensions, to demonstrate how arithmetic works with arrays.

5. Show how to clear the dimensions of an array.
```{r}
#p24, 5.9

nsnew <-ns
dim(nsnew)<-c(1,8) #adding dimensions
c(nsnew)

dim(nsnew) = NULL #deletes that attribute

```

### Chapter 6

1. Demonstrate the different ways of subsetting a list.
```{r}
$ []

```

2. What is the difference between giving multiple lists to `list()` versus `c()`?

3. How do data frames in R relate to lists?

4. What is the purpose of `attach()`?

### Chapter 7

1. Demonstrate the different approaches to reading in data from an external file.

### Chapter 8

1. Demonstrate the use of distribution functions to simulate data and calculate probabilities and quantiles.

2. Simulate data from a t distribution and check their normality.

### Chapter 9

1. What is the difference between `if() else` and `ifelse()`?

2. Demonstrate the use of a `for()` loop.