---
title: "EDPS 845 Assignment 1"
author: "Tony Albano"
---

## Overview

Demonstrate your mastery of concepts and procedures covered in *An Introduction to R*, chapters 1 through 9. See Appendix A for examples.

Complete this assignment by responding to each question below using your own original code, in addition to answering any questions. Follow Google's [R coding conventions](https://google.github.io/styleguide/Rguide.xml), and use [R markdown](http://rmarkdown.rstudio.com/index.html) formatting for your responses, with example R code written as

```{r}
# Comment within some example code
```

Your file should have an Rmd extension, and you should check that it compiles to HTML using the *Knit* command in RStudio (this runs `knitr::knit()` on the file). Note that you can already *Knit* this file to see how it works.

## Exercises

### Chapter 1

1. Show that you can access the home page for the HTML R help menu, as well as help files for individual functions.

```{r, eval = FALSE}
#This will open html help
help.start()

#Both of the following will open the help file for the solve function 
help(solve)
?solve
```

2. Demonstrate correct and incorrect naming of R objects.

```{r}
#Best practice dictates NOT using periods, use _ instead
#names can include a number, but cannot begin with a number
#Every character matters and is case-sensitive

#correct
x1 <- c(1,2,3) #where x1 is the object
X1 <- c(2,3,4) #different from lowercase x1

#incorrect
# 2x <- c(3,4,5)
```

3. How do you list all available objects in the R workspace? And how do you remove individual objects?

```{r}
# to display the objects which are currently stored within R
objects()

# to remove individual objects
rm(x1,X1)
```


### Chapter 2

1. Create and index a vector using a logical, integer, negative integer, and character vector.

```{r}

v1 <- 1:30 #creating a vextor that includes the integers 1, 2,...29, 30

#logical vector index

v1logical <- v1[v1%%2==0] #vector "v1logical" where only the values that are included are those for which the index vector in brackets is TRUE. Here, this would give you only the even values. 

# positive integer index

v1first10 <- v1[1:10] #vector that selects the first ten values of vector v1

#negative integer index

v1last10 <- v1[-(1:20)] #vector that excludes values 1-20 of vector v1

#character vector index

faces <- c(12, 11, 36, 24) #creates a vector of those values
names(faces) <- c("happy", "angry", "neutral", "surprised") #assigns names to identify each component of a vector
stim_list1 <- faces[c("neutral", "surprised")] #now I can create another vector comprised of the numeric mode values, without having to remember the specific numeric values because they have names

```

2. Create a simple data frame containing a numeric variable, character variable, and factor with at least 3 levels. Use the data frame to demonstrate how R codes and handles missing data. 

### Chapter 3

1. Demonstrate how to check and change the modes and attributes of R objects.

```{r}

mode(v1) #returns what the mode is; in this case, "numeric"
length (v1) #returns the length; in this case, 30
attributes(faces) #returns additional, non-instrinsic attributes, like names

as.character(v1) #changes all numeric values in vector v1 to character mode
as.integer(v1) #changes all characters in vector v1 back to integer mode

v2 <- numeric() #creates empty vector of mode numeric
length (v2) <-20 #makes the length 20, all missing values (NA)
v2[5] <- 5.1 #makes the 5th value 5.1

attr(v2, "dim") <- c(5,4) #changes the dimensions to a 5-by-4 matrix

```


### Chapter 4

1. Complete the exercise on page 17, where you calculate 95% confidence intervals for state mean incomes.
```{r}

state <- c("tas", "sa", "qld", "nsw", "nsw", "nt", "wa", "wa","qld", "vic", "nsw", "vic", "qld", "qld", "sa", "tas","sa", "nt", "wa", "vic", "qld", "nsw", "nsw", "wa","sa", "act", "nsw", "vic", "vic", "act")

statef <- factor(state) 

incomes <- c(60, 49, 40, 61, 64, 60, 59, 54, 62, 69, 70, 42, 56,61, 61, 61, 58, 51, 48, 65, 49, 49, 41, 48, 52, 46,59, 46, 58, 43)

incmeans <- tapply(incomes, statef, mean) #apply the mean() function to each group of income by levels of statef

stderr <- function(x) sqrt(var(x)/length(x)) #creates a funciton to calculate standard error

incster <- tapply (incomes, statef, stderr) #apply the newly created stderr function to each group of income by levels of statef

ns <- table(statef) #get n for each state/providence

qt (.025, ns-1) #returns CI with lower tail included for each n

```


2. In what order are the levels of ordered factors stored in?
```{r}
#They are stored in alphabetical order, or in the order they were specified to factor if they were specified explicitly

ex1 <- c("blue", "red", "green")
ex1f <- factor (ex1) #alphabetical


ex1f2 <- factor(ex1, levels=c("blue", "red", "green"))
levels(ex1f2) #specified order

```

### Chapter 5

1. Create an array with dimensions `c(3, 6, 2)` and use it to demonstrate array indexing.
```{r}
a1 <- array(1:36, dim=c(3,6,2))
a1

a1[2,4,1] #finds an element at a specific dimension

```

2. Create a two dimensional array of 100 integers sampled with replacement from the vector 1:20, and use an index array to set the diagonal values to 0.
```{r}
a2 <- array(rep(1:20,length.out=20), dim = c(10,10))
a2

ad <- array(c(1:10, 1:10), dim=c(10,2))
a2[ad] <- 0
a2
```

3. Construct the same array using the `array()` function and using a vector with the `dim()` set. Prove they are the same.

```{r}
#I did it that way in 5.2, so...?
```


4. Use your array from 5.1, and another with the same dimensions, to demonstrate how arithmetic works with arrays.
```{r}
a3 <- array(2:37, dim=c(3,6,2))
a1and3 <- 2*a1*a3+1
a1and3
```

5. Show how to clear the dimensions of an array.
```{r}
vec <- as.vector(a2)
vec
```

### Chapter 6

1. Demonstrate the different ways of subsetting a list.
```{r}
Lst <- list(name="Fred", wife="Mary", no.children=3, child.ages=c(4,7,9))

Lst$name #Fred
Lst[["wife"]] #Mary
Lst[[4]][2] #7
```

2. What is the difference between giving multiple lists to `list()` versus `c()`?
```{r}
#giving mulitple lists to 'list()'

#creating some lists to work with
Lst1 <- list(name="Catie",age=21,pets=1)
Lst2 <- list(name="Bob",age=23,pets=6)
Lst3 <- list(name="Mary",age=24,pets=0)

Lst4 <- list(Lst1,Lst2,Lst3) #using list()
Lst5 <- c(Lst1,Lst2,Lst3) #using c()

#when you compare the printouts of both of these, you can see that the c() combined lists do not have double-bracket operators for each of the 3 lists that were combined.
Lst4 
Lst5 

Lst4[[2]] [2] #to get Bob's age, it returns 23
Lst5[[2]] [2] #NA, can't do it!

#list() can combine lists that have elements of different types
```

3. How do data frames in R relate to lists?
```{r}
#A list or a matrix which satisfies the restrictions of a data frame can be coerced into a data frame using the function as.data.frame()
```

4. What is the purpose of `attach()`?
```{r, eval = FALSE}
#the function attach() makes data available to the R search path by taking the list or data frame as its argument

summary(women$age) #refers to variable 'age' in this example imaginary data frame
attach(women)
summary(age) #unil you detach(), you don't have to write women$ to refer to each variable in that data frame
```

### Chapter 7

1. Demonstrate the different approaches to reading in data from an external file.
```{r, eval = FALSE}
#input file form without row labels and with the first line as the column labels
HousePrice <- read.table("houses.data", header=TRUE) #houses.data is the file and header = true specifies that the first line is the column labels

inp <- scan("input.dat", list(id="", x=0, y=0)) #creates the object inp; input.dat is the file; the dummy list has named components

data(package="rpart") #access data from a particular package
```

### Chapter 8

1. Demonstrate the use of distribution functions to simulate data and calculate probabilities and quantiles.
```{r, eval = FALSE}
sim <- rnorm(100,0,1) #simulate a normal distribution of 100 characters with mean 0 and sd 1. 

2*pt(-2.43, df = 13) #2-tailed p-value for t distribution

qf(0.01, 2, 7, lower.tail = FALSE) #upper 1% point for an F(2, 7) distribution

```

2. Simulate data from a t distribution and check their normality.
```{r}
simt <- rt(300,299) #simulate t-dist
stem(simt) #examine stem-and-leaf plot
hist(simt) #examine histogram
plot(ecdf(simt), do.points=FALSE, verticals=TRUE) #plot the distribution function
shapiro.test(simt) #Shapiro-Wilk test
ks.test(simt, "pnorm", mean = mean(simt), sd = sqrt(var(simt))) #One-sample Kolmogorov-Smirnov test
```

### Chapter 9

1. What is the difference between `if() else` and `ifelse()`?
```{r}
#ifelse(test, what to return if test is true, what to return if test is false) is a vectorized version of the if/else concept that applies the test to each item of a vector and returns the true or false value for each item
ex1 <- c(1:12)
ifelse(ex1>=7,"upper","lower")

#if() else evaluates a condition once, unless part of a for loop

if (6<7) "RIGHT!" else "NO WAY!"
```

2. Demonstrate the use of a `for()` loop.
```{r}
#To get the square of each of the integers 1-5
num1 = c(1:5)
num1_sq = NULL

for (i in 1:5) {
num1_sq [i] = num1 [i]^2
}
print(num1_sq)
```

