---
title: "EDPS 845 Assignment 2"
author: "Catherine Brown"
date: "May 17, 2016"
output: html_document
---

## Data Structures

### Vectors

1. What are the six types of atomic vector? How does a list differ from an atomic vector?

All elements of an atomic vector must be the same type, whereas the elements of a list can have different types.
```{r}

#numeric/double
numvar <- c(1, 2.2, 3)
#integer -- L suffix gives you an integer thather than a numeric 
intvar <- c(1L, 2L, 3L)
#complex

#logical
logvar <- c(TRUE,F)
#character
chrvar <- c("string", "other string")
#raw -- creates vector of specified length where each value is 0
rawvar <- raw(2)
```


2. What makes is.vector() and is.numeric() fundamentally different to is.list() and is.character()?

is.vector() tests only if the object is a vector with no attributes apart from names, not whether something is a vector or not. Similarly, is.numeric returns TRUE for both integer and numeric/double vectors, not specifically numeric/double. It checks for "number-ness."

On the other hand, is.vector and is.numeric test if something is that type. 

3. Test your knowledge of vector coercion rules by predicting the output of the following uses of c():
```{r}
c(1, FALSE) 
  # numeric
c("a", 1) 
  # character
c(list(1), "a") 
  # list
c(TRUE, 1L)
  # numeric
```

4. Why do you need to use unlist() to convert a list to an atomic vector? Why doesn’t as.vector() work?

It won't work because lists are already vectors, and as.vector would just remove its list-related attributes. You have to unlist to make it a not-list vector and keep the attributes.

5. Why is 1 == "1" true? Why is -1 < FALSE true? Why is "one" < 2 false?

If the two arguments are different types of vectors, one is coerced into the type of the other, with the order of precedence being character, complex, numeric, integer, logial, and raw. 

1 == "1" becomes "1" == "1" where both are character type. Since they're the 
  same exact character, the statement is true.
-1 < FALSE becomes -1 < 0 where both are numeric type and the statement is true
"one" < 2 becomes "one" < "two" where both are characters and cannot be bigger
  or smaller than eachother, so the statement is false.

6. Why is the default missing value, NA, a logical vector? What’s special about logical vectors? (Hint: think about c(FALSE, NA_character_).)

Since logicals are almost at the bottom of the vector-coercion food chain (only ahead of raw), they don't coerce other vectors "down" to that type. This is good for NA, because a missing value that could coercively change the type of an entire vector would be problematic, especially since NA is common in data sets.

### Attributes and Factors

1. An early draft used this code to illustrate structure():

```{r}
structure(1:5, comment = "my attribute") #> [1] 1 2 3 4 5
```

But when you print that object you don’t see the comment attribute. Why? Is the attribute missing, or is there something else special about it? (Hint: try using help.)

R doesn't print the comment attribute, which is an exception to attributes in general. If you named the structure, then you could view the comment by using the following code:

```{r}

str <- structure(1:5, comment = "This is my structure.")
comment(str) <- "Just a comment."

```


2. What happens to a factor when you modify its levels?

It modifies the factor itself too. 
```{r}
f1 <- factor(letters)
f1 #prints a, b, c...
    
levels(f1) <- rev(levels(f1)) #reverses the levels
f1 #now prints z, w, x...
```

3. What does this code do? How do f2 and f3 differ from f1?

```{r}
#f1 created a vector where both the vector and the levels were in order

f2 <- rev(factor(letters)) #creates a reversed vector with levels that are not
  #reversed

f3 <- factor(letters, levels = rev(letters)) #creates a vector with only the 
  #levels reversed

```

### Matrices and Arrays

1. What does dim() return when applied to a vector?
NULL
```{r}
dim(f3)
```

2. If is.matrix(x) is TRUE, what will is.array(x) return?
TRUE. A matrix is an array.
```{r}
a <- matrix(1:6, ncol = 3, nrow = 2)
is.matrix(a)
is.array(a)
```

3. How would you describe the following three objects? What makes them different to 1:5?

```{r}
x1 <- array(1:5, c(1, 1, 5)) #1 row, 1 column, 5 deep
x2 <- array(1:5, c(1, 5, 1)) #1 row, 5 columns, 1 deep
x3 <- array(1:5, c(5, 1, 1)) #5 rows, 1 column, 1 deep
x4 <- array(1:5) #just has 1 row, no other specified dimensions

x1
x2
x3
x4
```

### Data Frames 

1. What attributes does a data frame possess?

Since it is a 2-dimensional structure consisting of equal-length vectors, a data frame shares properties of both the matrix and list. It has the attributes names (same as col.names), and row.names, and class. 

```{r}
dfm <- data.frame(x = 1:3, y = I(matrix(1:9, nrow = 3)))
attributes(dfm)
```

2. What does as.matrix() do when applied to a data frame with columns of different types?
```{r}
df <- data.frame(x = 1:3) #data frame where x is integer type
df$y <- list(1:2, 1:3, 1:4) #y is a list type
class(df$x)
class(df$y)
df
as.matrix(df) #the list y gets coerced into integer type
```

3. Can you have a data frame with 0 rows? What about 0 columns?
Yep.
```{r}
dfe <- data.frame()
dfe #returns "data frame with 0 columns and 0 rows"
```


## Subsetting

1. How would you randomly permute the columns of a data frame? (This is an important technique in random forests.) Can you simultaneously permute the rows and columns in one step?

2. How would you select a random sample of m rows from a data frame? What if the sample had to be contiguous (i.e., with an initial row, a final row, and every row in between)?

3. How could you put the columns in a data frame in alphabetical order?

