---
title: "EDPS 845 Assignment 2"
author: "Catherine Brown"
date: "May 20, 2016"
output: html_document
---

## Data Structures

### Vectors

1. What are the six types of atomic vector? How does a list differ from an atomic vector?

All elements of an atomic vector must be the same type, whereas the elements of a list can have different types.
```{r}

#numeric/double
numvar <- c(1, 2.2, 3)
#integer -- L suffix gives you an integer thather than a numeric 
intvar <- c(1L, 2L, 3L)
#complex...which I don't really understand 
#character
chrvar <- c("string", "other string")
#raw -- creates vector of specified length where each value is 0
rawvar <- raw(2)
```


2. What makes is.vector() and is.numeric() fundamentally different to is.list() and is.character()?

is.vector() tests only if the object is a vector with no attributes apart from names, not whether something is a vector or not. Similarly, is.numeric returns TRUE for both integer and numeric/double vectors, not specifically numeric/double. It checks for "number-ness."

On the other hand, is.vector and is.numeric test if something is that type. 

3. Test your knowledge of vector coercion rules by predicting the output of the following uses of c():
```{r}
c(1, FALSE) 
  # numeric
c("a", 1) 
  # character
c(list(1), "a") 
  # list
c(TRUE, 1L)
  # numeric
```

4. Why do you need to use unlist() to convert a list to an atomic vector? Why doesn’t as.vector() work?

It won't work because lists are already vectors, and as.vector would just remove its list-related attributes. You have to unlist to make it a not-list vector and keep the attributes.

5. Why is 1 == "1" true? Why is -1 < FALSE true? Why is "one" < 2 false?

If the two arguments are different types of vectors, one is coerced into the type of the other, with the order of precedence being character, complex, numeric, integer, logial, and raw. 

1 == "1" becomes "1" == "1" where both are character type. Since they're the 
  same exact character, the statement is true.
-1 < FALSE becomes -1 < 0 where both are numeric type and the statement is true
"one" < 2 becomes "one" < "two" where both are characters and cannot be bigger
  or smaller than eachother, so the statement is false.

6. Why is the default missing value, NA, a logical vector? What’s special about logical vectors? (Hint: think about c(FALSE, NA_character_).)

Since logicals are almost at the bottom of the vector-coercion food chain (only ahead of raw), they don't coerce other vectors "down" to that type. This is good for NA, because a missing value that could coercively change the type of an entire vector would be problematic, especially since NA is common in data sets.

### Attributes and Factors

1. An early draft used this code to illustrate structure():

```{r}
structure(1:5, comment = "my attribute") #> [1] 1 2 3 4 5
```

But when you print that object you don’t see the comment attribute. Why? Is the attribute missing, or is there something else special about it? (Hint: try using help.)

R doesn't print the comment attribute, which is an exception to attributes in general. If you named the structure, then you could view the comment by using the following code:

```{r}

str <- structure(1:5, comment = "This is my structure.")
comment(str) <- "Just a comment."

```


2. What happens to a factor when you modify its levels?

It modifies the factor itself too. 
```{r}
f1 <- factor(letters)
f1 #prints a, b, c...
    
levels(f1) <- rev(levels(f1)) #reverses the levels
f1 #now prints z, w, x...
```

3. What does this code do? How do f2 and f3 differ from f1?

```{r}
#f1 created a vector where both the vector and the levels were in order

f2 <- rev(factor(letters)) #creates a reversed vector with levels that are not
  #reversed

f3 <- factor(letters, levels = rev(letters)) #creates a vector with only the 
  #levels reversed

```

### Matrices and Arrays

1. What does dim() return when applied to a vector?
NULL
```{r}
dim(f3)
```

2. If is.matrix(x) is TRUE, what will is.array(x) return?
TRUE. A matrix is an array.
```{r}
a <- matrix(1:6, ncol = 3, nrow = 2)
is.matrix(a)
is.array(a)
```

3. How would you describe the following three objects? What makes them different to 1:5?

```{r}
x1 <- array(1:5, c(1, 1, 5)) #1 row, 1 column, 5 deep
x2 <- array(1:5, c(1, 5, 1)) #1 row, 5 columns, 1 deep
x3 <- array(1:5, c(5, 1, 1)) #5 rows, 1 column, 1 deep
x4 <- array(1:5) #just has 1 row, no other specified dimensions

x1
x2
x3
x4
```

### Data Frames 

1. What attributes does a data frame possess?

Since it is a 2-dimensional structure consisting of equal-length vectors, a data frame shares properties of both the matrix and list. It has the attributes names (same as col.names), and row.names, and class. 

```{r}
dfm <- data.frame(x = 1:3, y = I(matrix(1:9, nrow = 3)))
attributes(dfm)
```

2. What does as.matrix() do when applied to a data frame with columns of different types?
```{r}
df <- data.frame(x = 1:3) #data frame where x is integer type
df$y <- list(1:2, 1:3, 1:4) #y is a list type
class(df$x)
class(df$y)
df
as.matrix(df) #the list y gets coerced into integer type
```

3. Can you have a data frame with 0 rows? What about 0 columns?
Yep.
```{r}
dfe <- data.frame()
dfe #returns "data frame with 0 columns and 0 rows"
```


## Subsetting

### Data Types

1. Fix each of the following common data frame subsetting errors:
```{r}
#mtcars[mtcars$cyl = 4, ] 
mtcars[mtcars$cly == 4, ] # = is used for assignment, whereas == tests for exact equality

#mtcars[-1:4, ] you can't mix positive and negative integers in a single subset
mtcars[-c(1:4), ] # This will omit elements at positions 1 through 4

#mtcars[mtcars$cyl <= 5] does not have columns selected
mtcars [mtcars$cyl <= 5, ]

#mtcars[mtcars$cyl == 4 | 6, ]
mtcars[mtcars$cyl == 4 | mtcars$cyl == 6, ] #if you want the cars with 4 or 6 cyl,
  #you should specify mtcars$cyl before both 4 and 6, otherwise the second part
  #of the "or" statement means including all rows with values for the 6th column
```

2. Why does x <- 1:5; x[NA] yield five missing values? (Hint: why is it different from x[NA_real_]?)
NA is logical vector, and logical vectors get recycled when they are shorter than the vector being subsetted. When NA is real, it doesnt get recycled. And a missing value on the index returns a missing value on the output.
```{r}
x <- 1:5
x [NA] #yields 5 NA
x[NA_real_] # real coerces NA to double/numeric type, so it doesn't get recycled
```

3. What does upper.tri() return? How does subsetting a matrix with it work? Do we need any additional subsetting rules to describe its behaviour?

upper.tri() returns a matrix of logicals the same size of a given matrix with entries TRUE in the upper triangle.
upper.tri(x, diag = FALSE) 
x is a matrix
diag is a logical that excludes the diagonal

However, in the following code, one should note that it returns an atomic vector, not a matrix. The upper triangle's values are listed in column order.  

```{r}
x <- outer(1:5, 1:5, FUN = "*")
x[upper.tri(x)]
```

4. Why does mtcars[1:20] return an error? How does it differ from the similar mtcars[1:20, ]?

As Hadley Wickham describes, "Data frames possess the characteristics of both lists and matrices: if you subset with a single vector, they behave like lists; if you subset with two vectors, they behave like matrices."

In the first example, the subset only includes 1 vector, so you're asking for columns 1-20 as if it were a list. However, that many columns don't exist. In the second example, the subset includes 2 vectors, and you're asking for rows 1:20 with all columns included as if it were a matrix.

5. Implement your own function that extracts the diagonal entries from a matrix (it should behave like diag(x) where x is a matrix).
```{r}
matr <- matrix(1:25, nrow=5) #made a matrix
diag(matr) #it could have been this easy, but no...

dfun <- function(x) {
  stopifnot(nrow(x) == ncol(x), is.matrix(x)) #has to be a square and matrix
  n <- ncol(x) #alternativey, could be nrow(x), since they're the same
  vec <- vector() #create empty vector
  for(i in 1:n){ #for each value along the number of columns...
    vec [i] <- x[i,i] #put into our empty vector the subset of the matrix x 
    #which is in row i, column i (the diagonal)
  }
  return(vec) #show me the vector
}  
dfun(matr)
```

6. What does df[is.na(df)] <- 0 do? How does it work?
It returns a logical matrix that has the same dimensions as the original data frame, but where all NA values are 0. Every value in the data fram goes through a logical test of being TRUE or FALSE with regard to is.na. For those which are TRUE, the df[] <- 0 applies.
```{r}
df <- data.frame(x = c(1, 2, NA), y = c(NA, 4, 5))
df
df[is.na(df)] <- 0
df
```

### Subsetting Operators

1. Given a linear model, e.g., mod <- lm(mpg ~ wt, data = mtcars), extract the residual degrees of freedom. Extract the R squared from the model summary (summary(mod))
```{r}
lmod <- lm(mpg ~ wt, data = mtcars)
(df_lmod <- lmod[["df.residual"]])
```

###Subsetting and Assignment; Applications

1. How would you randomly permute the columns of a data frame? (This is an important technique in random forests.) Can you simultaneously permute the rows and columns in one step?
```{r}
df <- data.frame(x = 1:6, y = 6:1, z = letters[1:6]) #create data frame
df

df[,sample(ncol(df))] #randomly reorder columns
df[sample(nrow(df)),sample(ncol(df))] #randomly reorder rows and columns
```

2. How would you select a random sample of m rows from a data frame? What if the sample had to be contiguous (i.e., with an initial row, a final row, and every row in between)?
```{r}
df <- data.frame(x = 1:26, y = 26:1, z = letters) #create data frame
m = 18
df[sample(nrow(df), size = m), ] #returns a subset of m random rows from df
df[sample(nrow(df), size = 1) + (0:(m - 1)), ] #returns 1 random row from df,
  #as well as the 0-(n-1) rows after that
```

3. How could you put the columns in a data frame in alphabetical order?
```{r}
mtcars[ ,order(names(mtcars))] #the comma is important-- rows, columns
```

## Functions

### Function Components

1. What function allows you to tell if an object is a function? What function allows you to tell if a function is a primitive function?
```{r}
is.function(mtcars) #FALSE
is.function(dfun) #TRUE
is.primitive(dfun) #FALSE
is.primitive(sqrt) #TRUE, internally implemented function of R
```

2. This code makes a list of all functions in the base package.
```{r}
objs <- mget(ls("package:base"), inherits = TRUE)
funcs <- Filter(is.function, objs)
```
Use it to answer the following questions:
a. Which base function has the most arguments?
```{r}
argn <- sapply(funs, function(x) length(formals(x))) #goes through all of funs
  #and applies function x to get the length of all formal arguments
which.max(argn) #determines the location of the maximum of the vector argn. 
```

b. How many base functions have no arguments? What’s special about those functions?
They're special because they're all primitive functions.
```{r}
length(which(argn == 0)) #get the length of a vector which includes all argn == 0
```

c. How could you adapt the code to find all primitive functions?
```{r}
names(Filter(is.primitive, objs)) #extracts all elements of objs for which the
  #logical argument is true, and gets the names
```

3. What are the three important components of a function?
As Wickham states:
* the body(), the code inside the function.
* the formals(), the list of arguments which controls how you can call the function.
* the environment(), the “map” of the location of the function’s variables.
```{r}
body(dfun)
formals(dfun)
environment(dfun)
```

4. When does printing a function not show what environment it was created in?
It doesn't show the environment if it was created in the global environment.
```{r}
environment(sqrt)
```

## Functions

### Lexical Scooping

1. What does the following code return? Why? What does each of the three c’s mean?
```{r}
c <- 10
c(c = c)
```

The first c <- 10 is an object, the 2nd c() is a concatenate function, and the 3rd c = c is an item named c with the value of the object c, which is 10.

```{r}
#Another way to get the same output:
d <- 10
c(c = d)
```

2. What are the four principles that govern how R looks for values?
Per Wickham, 
"There are four basic principles behind R’s implementation of lexical scoping:
* name masking
* functions vs. variables
* a fresh start
* dynamic lookup"
```{r}
#Name masking: uses the most recent assignment
x <- 2
g <- function() {
  y <- 1
  c(x, y) #R figures out that x's assignment from above
}
g()

#Functions vs. variables: the same lexical scoping principles apply for both
#However, if you confusingly give the same name to a funciton and a variable,
  #R will ignore non-funcitons when it's clear you want the function
n <- function(x) x / 2
o <- function() {
  n <- 10
  n(n)
  }
o()

#A fresh start: every time a function is called, a new environment is created 
j <- function() {
  if (!exists("a")) {
    a <- 1
  } else {
    a <- a + 1
  }
  print(a)
}
j()

#Dynamic Lookup:R looks for values when the function is run, not when it’s created. 
  #This means that the output of a function can be different depending on objects  
  #outside its environment:
f <- function() x
x <- 15
f()
x <- 20
f() #so if you weren't keeping track of the value assigned to x, you could get 
  #confused down the line when you try to use f()

rm(x, g, n, o, j)
```

3. What does the following function return? Make a prediction before running the code yourself.
Start at the "deepest" level, x^2 = 100. 
Pop out a level, 100 + 1 = 101.
Pop out one more level, 101 * 2 = 202.
```{r}
f <- function(x) {
  f <- function(x) {
    f <- function(x) {
      x ^ 2
    }
    f(x) + 1
  }
  f(x) * 2
}
f(10)
```

### Function Arguments

1. Clarify the following list of odd function calls:
```{r}
x <- sample(replace = TRUE, 20, x = c(1:10, NA))
#sample(x, size, replace = FALSE, prob = NULL)
x <- sample(c(1:10, NA), 20, replace = TRUE)

y <- runif(min = 0, max = 1, 20)
#runif(n, min = 0, max = 1)
y <- runif(20, 0, 1)

cor(m = "k", y = y, u = "p", x = x) 
#cor(x, y = NULL, use = "everything", method = c("pearson", "kendall", "spearman"))
cor(x, y, use = "pairwise", method = "kendall")
```

2. What does this function return? Why? Which principle does it illustrate?
It demonstrates a fresh start because it creates a new environment when f1() is called, and it evaluates y <- 1 instead of the more recent y = 0.
```{r}
f1 <- function(x = {y <- 1; 2}, y = 0) {
  x + y #where x = 2 and y = 1
  }
f1()
```

3. What does this function return? Why? Which principle does it illustrate?
It illustrates dynamic lookup because x doesn't exist outside the function, but R doesn't look for it until you use it.
```{r}
f2 <- function(x = z) {
  z <- 100
  x 
  }
f2()
```

### Special Calls

1. Create a list of all the replacement functions found in the base package. Which ones are primitive functions?
```{r}
objs <- mget(ls("package:base"), inherits = TRUE)
funs <- Filter(is.function, objs)
repfs <- funs[grep("<-", funs)] #subset of funs that searches funs for "<-"
names(repfs) # List of all replacement functions in the base package
primfs <- Filter(is.primitive, repfs) 
names (primfs)# List of all primitive replacement functions in base
```

2. What are valid names for user-created infix functions?

Anything, so long as it starts and ends with % and doesn't have nothing, /, *, o, x, or in in between, as these are already predefined infix operators.

3. Create an infix xor() operator.
I need some help understanding the point of user-created infixes.

```{r}
`%xor%` <- function(a,b)  a != b
`%xor%` (1, 2)
```

4. Create infix versions of the set functions intersect(), union(), and setdiff().
```{r}
`%intr%` <- function(a,b) intersect(a, b)
`%union%` <- function (a,b) union(a,b)
`%setdf%` <- function (a,b) setdiff(a, b)
  
`%inter%` (1:10, 5:15)
`%union%` (1:5, 5:10)
`%setdf%` (1:10, 5:10)
```

5. Create a replacement function that modifies a random location in a vector.
```{r}
rmod <- function(x) {
  location <- sample(1:length(x), size = 1) #choose 1 item from a vector the
    #size of x
}

chunk <- 1:5 
rmod(chunk) <- 0 #that single item becomes 0
chunk
```

### Return Values

1. How does the chdir parameter of source() compare to in_dir()? Why might you prefer one approach to the other?

When chdir = TRUE in source(), the R working directory is temporarily changed to the location of the source code (the first element of the source function). With in_dir(), on the other hand, you can set the directory to any location, even if it is not the location of the source code.

2. What function undoes the action of library()? How do you save and restore the values of options() and par()?

```{r, eval=FALSE}
library(devtools)
detach(devtools)
```

Save the values of options() and par() as objects, then define them as the options() and par() in the new.

```{r, eval = FALSE}
othr_opt <- options ()
othr_par <- par()

options(othr_opt)
par(othr_par)

#Would this work?
othr_opt <- options ()
othr_par <- par()
on.exit(options(othr_opt))
on.exit(par(othr_par))
```

3. Write a function that opens a graphics device, runs the supplied code, and closes the graphics device (always, regardless of whether or not the plotting code worked).
```{r}
plotf <- function() {
  plot(x, y)
  on.exit(dev.off()) #shuts down plotting device
}
```

4. We can use on.exit() to implement a simple version of capture.output().
```{r}
capture.output2 <- function(code) {
  temp <- tempfile()
  on.exit(file.remove(temp), add = TRUE)

  sink(temp)
  on.exit(sink(), add = TRUE)

  force(code)
  readLines(temp)
}
capture.output2(cat("a", "b", "c", sep = "\n"))

capture.output
```

Compare capture.output() to capture.output2(). How do the functions differ? What features have I removed to make the key ideas easier to see? How have I rewritten the key ideas to be easier to understand?

capture.output2 is certainly simpler and easier to understand, but it seems to accomplish the same end as capture.output. 